---
title: "R Notebook"
author: "Ousam Salah"
data: 01/12/2022
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

### Case Study

This R notebook is aiming to answer business tasks and deliver useful insights and high recommendations to reveal more opportunities for growth.

### Business Plan

Bellabeat, a high-tech company that manufactures health-focused smart products. Our objective is to analyze smart devices usage data in order to gain insight about how consumers use Bellabeat products, and how theses insights can identify potential opportunities for growth and recommendations for the Bellabeat marketing strategy improvement.

-   What are some trends in smart device usage ?
-   How could these trends apply to Bellabeat customers ?
-   How could these trends help influence Bellabeat marketing strategy ?

After we identified the business task , we should now prepare and load the data needed for the analysis , **the data we are going to use is an open dataset from kaggle which can be found [here](https://www.kaggle.com/datasets/arashnic/fitbit)**. the data source contains many different CSV files, we are going choose a certain files to work with.

```{r}
##load libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
```

The analysis will focus on daily inputs as it is better for discovering health benefits trends.

```{r}
# Import all all relevant csv files as R data frames
dailyActivity <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
dailySleep <- read_csv("Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
weightInfo <- read_csv("Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv")

# Preview the data frames
head(dailyActivity)
head(dailySleep)
head(weightInfo)

```

All three data frames need some formatting to be usable and can be merged with the Id and Date keys.

```{r}
# Cleaning and transforming the datas

#dailyActivity dataset
dailyActivity <- dailyActivity %>%
  rename(Date = ActivityDate) %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%y"))

#dailySleep dataset
dailySleep <- dailySleep %>% 
  rename(Date = SleepDay) %>% 
  mutate(Date = as.Date(Date, format = "%m/%d/%y"))

#weightInfo dataset
weightInfo <- weightInfo %>% 
  select(-LogId) %>% 
  mutate(Date = as.Date(Date, format = "%m/%d/%y")) %>% 
  mutate(IsManualReport = as.factor(IsManualReport))

```

```{r}
#Merging all datas into one dataframe to simplify analysis

fitInfo <- merge(dailyActivity, dailySleep, by = c('Id', 'Date'), all = TRUE)
fitInfo <- merge(fitInfo, weightInfo, by = c('Id','Date'), all = TRUE)
head(fitInfo)
```

```{r}
# Let's determine the percentage of empty observation by variable
(colMeans(is.na(fitInfo)))*100
```

We can see that BMI, Manual Report, Fat Weights have more than 90% NA values. There is also duplicate columns so we can remove them from the data frame.

```{r}
fitInfo <- fitInfo %>% select(-c(TrackerDistance, LoggedActivitiesDistance, TotalSleepRecords, WeightPounds, WeightKg, Fat, BMI, IsManualReport))
```

Lets have a summary of the data frame

```{r}
str(fitInfo)
```

```{r}
# explore numerical varialbes per category
fitInfo %>% 
  select(-c(Id,Date))%>%
  summary()
```

**Insights from this snapshot:**

-   Average sedentary time is 17h30 out 24h (Not good).

-   The majority of the participants are lightly active.

-   On the average, participants sleep for 7 hours (with 530 missing observation).

-   Average total steps per day are 7652 which is around 5.5km and little below standards for having health benefits according to medical research.

```{r}
# Create a scater plot of totals steps and calories for all users
ggplot(data=fitInfo, aes(x=TotalSteps, y=Calories)) + 
  geom_point(color='deepskyblue2') + 
  geom_smooth() + 
  labs(title="Calories burned by steps")
```

There is positive correlation of 0.65 between the two variables plots, which is natural because activity is linked to calories burned in ours body.

```{r}
#Box plots of the steps taken by users each day of the study
fitInfo %>% 
  mutate(weekdays = weekdays(Date)) %>% 
  select(weekdays, TotalSteps) %>% 
  mutate(weekdays = factor(weekdays, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'))) %>% 
  drop_na() %>% 
  ggplot(aes(weekdays, TotalSteps, fill = weekdays)) +
  geom_boxplot()+
  labs(title = "Steps taken by days of the week")+
  theme(legend.position="none")
```

There is not any particular trend appearing as average total steps stays between (5 000-10 000) expect for some outliers between Wednesday and Friday, meaning that days of the week are not factors of activity.

```{r}
#plotting threes graphs to check sleeps in users
sleepTime <- fitInfo$TotalMinutesAsleep
bedTime <- fitInfo$TotalTimeInBed
par(mfrow=c(1,3))
hist(sleepTime, col = "#ff3993")
hist(bedTime, col = "#3399cc")
sleepQuality = sleepTime/bedTime
hist(sleepQuality, col = "#66CC99")

```

We can see that most of the sample have good sleep to time in bed ration meaning they don't have insomnia. Let's some factors that can explain why.

```{r}
fitInfo %>% 
  select(TotalMinutesAsleep, TotalSteps) %>% 
  mutate(sleep_quality = ifelse(TotalMinutesAsleep <= 420, 'Less than 7h',
                                ifelse(TotalMinutesAsleep <= 540, '7h to 9h', 
                                       'More than 9h'))) %>% 
  mutate(active_level = ifelse(TotalSteps >= 15000,'More than 15,000 steps',
                               ifelse(TotalSteps >= 10000,'10,000 to 14,999 steps',
                                      ifelse(TotalSteps >= 5000, '5,000 to 9,999 steps',
                                             'Less than 4,999 steps')))) %>% 
  select(-c(TotalMinutesAsleep, TotalSteps)) %>% 
  drop_na() %>% 
  group_by(sleep_quality, active_level) %>% 
  summarise(counts = n()) %>% 
  mutate(active_level = factor(active_level, 
                               levels = c('Less than 4,999 steps',
                                          '5,000 to 9,999 steps',
                                          '10,000 to 14,999 steps',
                                          'More than 15,000 steps'))) %>% 
  mutate(sleep_quality = factor(sleep_quality, 
                                levels = c('Less than 7h','7h to 9h',
                                           'More than 9h'))) %>% 
  ggplot(aes(x = sleep_quality, 
             y = counts, 
             fill = sleep_quality)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values=c("tan1", "#66CC99", "lightcoral")) +
  facet_wrap(~active_level, nrow = 1) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(strip.text = element_text(colour = 'black', size = 8)) +
  theme(strip.background = element_rect(fill = "beige", color = 'black'))+
  labs(
    title = "Sleep quality by steps",
    x = "Sleep quality",
    y = "Count"
  )
```

We can see that the sample with the best sleep time quality take [5000 - 10000] steps per day.

### Summarizing recommendations for the business

-   Calories are burnt by the steps taken daily. Based on users objectives, Bellabeat could recommend a minimum number of steps for users to take to encourage them to achieve their goals.

-   In order to create a habit of exercising every day, Bellabeat could send a notification at a specific time for the user to remain consistent throughout the week.

-   The data shows that in order to sleep better the best type of exercise is light to moderate (less than 10,000 steps). Bellabeat could recommend this level of exercise for people who want to simply maintain a healthy lifestyle.
